%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                      %
% GLE - Graphics Layout Engine <http://www.gle-graphics.org/>          %
%                                                                      %
% Modified BSD License                                                 %
%                                                                      %
% Copyright (C) 2009 GLE.                                              %
%                                                                      %
% Redistribution and use in source and binary forms, with or without   %
% modification, are permitted provided that the following conditions   %
% are met:                                                             %
%                                                                      %
%    1. Redistributions of source code must retain the above copyright %
% notice, this list of conditions and the following disclaimer.        %
%                                                                      %
%    2. Redistributions in binary form must reproduce the above        %
% copyright notice, this list of conditions and the following          %
% disclaimer in the documentation and/or other materials provided with %
% the distribution.                                                    %
%                                                                      %
%    3. The name of the author may not be used to endorse or promote   %
% products derived from this software without specific prior written   %
% permission.                                                          %
%                                                                      %
% THIS SOFTWARE IS PROVIDED BY THE AUTHOR "AS IS" AND ANY EXPRESS OR   %
% IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED       %
% WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE   %
% ARE DISCLAIMED. IN NO EVENT SHALL THE AUTHOR BE LIABLE FOR ANY       %
% DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL   %
% DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE    %
% GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS        %
% INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER %
% IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR      %
% OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN  %
% IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.                        %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%\chapter {GLE Utilities}
\section{Manip}
\index{Manip}
%
Manip is a data manipulation package.  It reads in a text
file of numbers and displays them like a spreadsheet.
You can then do simple operations on the columns and write
them out in any format you like.

\subsection{Usage}
\index{Manip!usage}
\begin{verbatim}
    MANIP infile.dat -recover -step -commands c.log -single -size x y
\end{verbatim}

\begin{commanddescription}
\item[{\sf --recover}]
\index{recover}
Manip logs everything you type to a file called \verb#MANIP_.J1#
When you use the {\sf -RECOVER} option on the manip command it then
reads keys from that file as if they were typed at the keyboard.

This will restore you to the 
point just before your pc crashed.
The last three journal files are stored (.j1 .j2 .j3)
simply copy the one you want to (.j1) to use it.

\item[{\sf --step}]
\index{step}
Used with recover,  press a space for each key you
want to read from the journal file,  press any other
key to stop reading the journal.

\item[{\sf --commands {\it filename.man}}]
\index{commands}    
This reads the commands in {\it filename.man} as if they
were typed at the keyboard.

\item[{\sf --single}]
\index{single}
This makes {\sf MANIP} use single precision arithmetic
and doesn't store strings at all, this enables
three times as much data in the same amount of memory

\item[{\sf --size {\it x y}}]
\index{size {\it x y}}
Sets the initial size of the spreadsheet.  Use this
with large datasets as it prevents the heap from
becoming fragmented and thus lets you use much
larger datasets.
\end{commanddescription}

\subsubsection{Range}
\index{Manip!Range}
Most manip commands accept a range as one or more of there
parameters.  A range is a rectangular section of your
spreadsheet.  A range can ether start with a 'c' or an
'r' and this will affect how the command operates.

If your spreadsheet has 5 columns and 10 rows then.
\begin{verbatim}
		c1 ==  c1c1r1r10  == 1,1  1,2  1,3  1,4  1,5  1,6 ...
		r1 ==  r1r1c1c5   == 1,1  2,1  3,1  4,1  5,1
		c1c2 ==c1c2r1r10  == 1,1  2,1  1,2  2,2  3,1  3,2 ...
		r1r2c3=r1r2c3c5   == 3,1  3,2  4,1  4,2  5,1  5,2
\end{verbatim}

\subsubsection{Arrows}
\index{Manip!Arrows}
The arrow keys normally move the data cursor, however if you are half
way thru typing a command then, the left and right arrow keys allow you
to edit the command.  Use the {\sf PAGE-UP} and {\sf PAGE-DOWN} keys to recall
your last command.

{\sf SHIFT} arrow keys will jump 7 cells at a time for fast movement.\\

Further help is available on the following toppics via
the HELP command e.g. "HELP COPY"

\subsection{Manip Primitives (a summary)}
%
%%{\sf xplaces {\it pos1 pos2 pos3 ...}}  \\
%
{\sf @{\it mycmds}}\\
{\sf Arrows}\\
{\sf BLANK}\\
{\sf CLEAR}\\
{\sf CLOSE}\\
{\sf COPY [{\it range}] [{\it range}] IF [{\it exp}]}\\
{\sf DATA [{\it range}]}\\
{\sf DELETE [{\it range}] IF [{\it exp}]}\\
{\sf EXIT {\it filename} [{\it range}] --TAB --SPACE --COMMA}\\
{\sf FIT {\it c3}}\\
{\sf Functions}\\
{\sf GENERATE [{\it pattern}] [{\it destination}]}\\
{\sf GOTO {\it x y}}\\
{\sf INSERT [C{\it n}] or [R{\it n}]}\\
{\sf LOAD {\it filename} [{\it range}] --0}\\
{\sf LOAD {\it filename} [{\it range}]}\\
{\sf LOGGING {\it mycmds.man}}\\
{\sf MOVE [{\it range}] [{\it range}] IF [{\it exp}]}\\
{\sf NEW}\\
{\sf PARSUM [{\it range1}] [{\it range2}]}\\
{\sf PROP [{\it range}] [{\it range}]}\\
{\sf QUIT}\\
{\sf Recover (recovering from power failure or crash)}\\
{\sf SAVE {\it filename} [{\it range}] --TAB --SPACE --COMMA}\\
{\sf SET SIZE {\it ncols nrows} }\\
{\sf SET BETWEEN "  "}\\
{\sf SET COLTYPE}\\ 
{\sf SET COLWIDTH}\\
{\sf SET NCOL {\it n}}\\
{\sf SET DPOINTS {\it n}}\\
{\sf SET DIGITS {\it n}}\\
{\sf SET WIDTH {\it n}}\\
{\sf SHELL}\\
{\sf SORT [{\it range}] on [{\it exp}]}\\
{\sf SUM [{\it range}]}\\
{\sf SWAP C{\it n}C{\it n} | R{\it n}R{\it n}}

\subsection{Manip Primitives (in detail)}
%
\begin{commanddescription}
%%\item[{\sf data {\it filename {\sf [}d1 d2 d3 ...{\sf ] [}d1=c1,c3{\sf ]} }
%%} ]
%% \index{data}
\item[{\sf COPY [{\it range1}] [{\it range2}]  if [{\it exp}]}]
\index{copy}

For copying a section to another section.
They do not have to be the same shape. The pointers to both rangers
are increased even if the number is not coppied  e.g.
\begin{verbatim} 
	"% COPY r4r2 r1r2"
	"% COPY c1c3r6r100  c6c8 if c1<c2"

	"% COPY C1 C2 IF C1<4"
	c1 	c2
	1       1
	2       2
	5       -
	3       3
	9       -
\end{verbatim}

\item[{\sf DELETE [{\it range}] IF [{\it exp}]}]
\index{delete}
For deleteing entire rows or columns. e.g.
\begin{verbatim}
	"% DELETE c1c3 IF r1>3.and.r2=0
	"% DELETE r1"
\end{verbatim}
Numbers are shuffled in from the right to take the place of the
deleted range.

\item[{\sf DATA [{\it range}]}]
\index{data}
Data entry mode is usefull for entering data.  After typing in
\verb#"% DATA c1c3"# or \verb#"% DATA C2"# you can then enter data and pressing
{\sf <cr>} will move you to the next valid data position.
In this mode text or numbers can be entered.
Press {\sf ESC} to get back to command mode.

\item[{\sf FIT {\it c3}}]
\index{fit}
\verb#"FIT C3"# will fit a least squares regression line to the data
in columns c3 and c4 (x values taken from c3) and print
out the results.

\item[{\sf EXIT}]
\index{exit}
EXIT saves the data in your input file spec and exits to DOS.  You
can optionally specify an output file as well. eg. 
\verb#"% EXIT myfile.dat"#

The command \verb#"EXIT myfile.dat c3c5r1r3"# will write out that range of
numbers to the file.

By default manip will write columns seperated by spaces.

The command \verb#"EXIT myfile.dat -TAB"# will put a single tab between
each column of numbers and \verb#"EXIT myfile.dat -COMMA"# will put a
comma and a space between each number.  (these two options are
usefull if your data file is very big and you don't want to waste
diskspace with the space characters.)  
Note: The settings stay in effect for future saves and exits.

You can make it line up the columns on the decimal point by typing in the
command. \verb#"SET DPOINTS 3"#

You change the width of each column or completely remove the spaces
between columns with the command. \verb#"SET WIDTH 10"#  (or set width 0)

You can change the number of significant digits displayed with
the command \verb#"SET DIGITS 4"#

\item[{\sf SAVE {\it myfile.dat}}]
\index{save}
Saves all or part of your data.
The command \verb#"SAVE myfile.dat c3c5r1r3"# will write out that range of
numbers to the file.

By default manip will write columns seperated by spaces.

The command \verb#"SAVE myfile.dat -TAB"# will put a single tab between
each column of numbers and \verb#"SAVE myfile.dat -COMMA"# will put a
comma and a space between each number.  (these two options are
usefull if your data file is very big and you don't want to waste
diskspace with the space characters).

Further options are the same like {\sf EXIT}  

\item[{\sf GOTO}]
\index{goto}
For moving the cursor directly to a point in your array.
e.g. \verb#"% GOTO x y"#

\item[{\sf CLEAR}]
\index{clear}
\verb#"% CLEAR C2C3"# Clears the given range of all values

\item[{\sf BLANK}]
\index{BLANK}
\verb#"% BLANK C2C3"# Clears the given range of all values

\item[{\sf NEW}]
\index{new}
Clears the spread sheet of all data and frees memory.

\item[{\sf INSERT}]
\index{insert}
Inserts a new column or row and shifts all others over.
e.g.\verb#"% INSERT c5" or "% INSERT r2"#.

\item[{\sf LOAD}]
\index{load}
Load data into columns. eg. \verb#"% LOAD filename"# loads all 
data into corresponding columns. \verb#"% LOAD filename c3"#
load first column of data into c3 etc.

\verb# "LOAD myfile.dat c3 -LIST"#
This commmand will load the the data into a single column or range
(even if it is several columns wide in the data file)

\item[{\sf MOVE [{\it range1}] [{\it range2}] if [{\it exp}]}]
\index{move}
For copying a section to another section. They do not have to be the 
same shape. The pointer to the destination is only increased if 
the line or column is coppied e.g.
\begin{verbatim}
	"% MOVE c1 c2c3"
	"% MOVE r4r2 r1r2"
	"% MOVE c1c3r6r100  c6c8 if c1<c2"

	"% MOVE C1 C2 IF C1<4"
	c1 	c2
	1       1
	2       2
	5       3
	3       -
	9       -

	(See COPY command)
\end{verbatim}

\item[{\sf SORT [{\it range}] ON [{\it exp}]}]
\index{sort}
Sort entire rows of the data based on the data in a particular column. e.g.
\begin{verbatim}
	"% SORT c8 on c9"
	"% SORT c1c8 on -c8"
	"% SORT c1c3 on c2  " 	!for sorting strings

\end{verbatim}
This command works out how to sort the column (or exp) specified
in the {\sf ON} part of the command.  It then does that operation to
the range specified. e.g. \verb#"SORT C1 ON C1"#  will sort column one.

Use the additional qualifier \verb#-STRINGS# if you want to sort a
column with strings in it. e.g. \verb#"sort c1 on c2 -strings"#

\item[{\sf SWAP}]
\index{swap}
Swap over two columns or rows. e.g.
\begin{verbatim}
 	"% SWAP c1c2"
	"% SWAP r3r1"
\end{verbatim}

\item[{\sf SET SIZE {\it ncols nrows}}]
\index{set size}
\verb#"SIZE 3 4"# Truncates the spreadsheet to 3 columns and 4 rows.  This
also sets the values to use for default ranges.

\item[{\sf SET BETWEEN " "}]
\index{set between}
\verb$"SET BETWEEN "##"$ Defines the string to be printed 
between each column of numbers when written to a file. 
This is normally set to a single space.

\item[{\sf SET COLWIDTH}]
\index{set colwidth}
Set the width of each column when displayed.
e.g. \verb#"% SET COLWIDTH 12"#

\item[{\sf SET COLTYPE [{\it n}] DECIMAL | EXP | BOTH | DPOINTS {\it n}}]
\index{set coltype}
This commands allows all or individual columns to be set to
different output types.  If colnumber is missing then that
setting is applied to all columns.

\verb#SET COLTYPE Ccolnumber TYPE#  Where {\sf TYPE} is one of:
\begin{verbatim}
	DECIMAL    produces 	123.456
	EXP	   produces	1.23456e02
	BOTH       produces whichever is more suitable
	DPOINTS n  produces a fixed number of decimal places.
 e.g.
	SET COLTYPE c2 DECIMAL
	SET COLTYPE c1 EXP
	SET COLTYPE c3 DPOINTS 4
\end{verbatim}
Would print out: \verb# 1.2e02 	  1.2    1.2000#
\begin{verbatim}
	SET COLTYPE EXP   (column number missed out)
\end{verbatim}
Would print out: \verb# 1.2e02    1.2e02  1.2e02#

\item[{\sf SET NCOL {\it n}}]
\index{set ncol}
 Set the number of columns to display.  e.g.\verb#"% SET NCOL 3"#

\item[{\sf SET DPOINTS {\it n}}]
\index{set dpoints}
Sets the number of decimal places to print.  This is used for
producing columns which line up on the decimal point.
e.g. with \verb#DPOINTS 3#.
\begin{verbatim}
	2.2   ->   2.200
	234   -> 234.000
\end{verbatim}
(See also {\sf SET COLTYPE}) 

\item[{\sf SET DIGITS {\it n}}]
\index{set digits}
Sets the number of significat digits to be displayed, e.g.
with \verb#DIGITS 3#.
\begin{verbatim}
	123456    becomes   123000
	0.12345   becomes   0.123
\end{verbatim}

\item[{\sf SET WIDTH {\it n}}]
\index{set width}
Sets the width of padding to use for the columns when they
are written to a file.
The columns usually one space wider than this setting as
the {\sf BETWEEN} string is usually set to one space by default.

\item[{\sf LOGGING}]
\index{logging}
For creating command files. e.g.
\begin{verbatim}
	"% LOG sin.man"
	"% c2=sin(c1)
	"% c3=c2+2
	"% close"
\end{verbatim}
Then type in \verb#"@sin"# to execute these commands.

\item[{\sf PROPAGATE [{\it source}] [{\it destination}]}]
\index{propagate}
This command has the same format as move. 
The difference is that the source is coppied as many times as 
possible to fill up the destination. 
e.g. \verb#"% PROP c1r1r7 c2"#

\item[{\sf SUM [{\it range}]}]
\index{sum}
Adds up all the numbers in a range and displays the total and average.
e.g. \verb#"% SUM C1C3"#

\item[{\sf PARSUM [{\it range1}] [{\it range2}]}]
\index{parsum}
Adds up one coloumn, putting the partial sum's into another coloumn. 
e.g. \verb# 1,2,3,4# becomes \verb# 1,3,6,10#.

\item[{\sf GENERATE [{\it pattern}] [{\it destination}]}]
\index{generate}
For generating a patter of data e.g. \verb# 1 1 2 2 5 5 1 1 2 2 5 5# etc.
\begin{verbatim}
	"% GEN 2(1,2,5)30 c4"	!1 1 2 2 5 5  repeated 30 times
	"% GEN (1:100:5)5 c1"	!1 to 100 step 5,  5 times
	"% GEN (1,2,*,3:5)5 c1"	!missing values included
\end{verbatim}

\item[Functions]
\index{Functions}
Calculations can be performed on rows or columns. eg \verb#"% C1=C2*3+R"# 
where "R" stands for row-number and C1 and C2 are columns. 
They can also be performed on {\sf ROWS}. eg 
\begin{verbatim}
	r1=sin(r2)+log10(c)
	c1 = cell(c+1,r)+cell(c+2,r)
	cell(1,3) = 33.3
	3+4*COS(PI/180)^(3+1/30)+C1+R
\end{verbatim}
 Valid operators and functions:
%%\begin{verbatim}
{\tt
\begin{center}
\begin{tabular}{ccccccc}
 $,$   & $+$  & $-$  & \verb#^#  & $*$  & $/$  & $<=$  \\
 $>=$  & $<>$ & $<$  & $>$  & $=$  & )AND(& )OR(  \\
 ABS(  & ATN( & COS( & EXP( & FIX( & INT( & LOG(  \\
LOG10( & SGN( & SIN( & SQR( & TAN( & NOT( & RND(   \\
 SQRT( & .NE. & .EQ. & .LT. & .GT. & .LE. & .GE. \\
 .NOT. & .AND.& .OR. &      &      &      &       \\
\end{tabular}
\end{center}
}
%%\end{verbatim}

\item[{\sf QUIT}]
\index{quit}
Abandon file.


\item[{\sf SHELL}]
\index{shell}
Gives access to DOS.
\end{commanddescription}
